---
title: "Project 2: Top songs, artists, and months on Spotify"
author: "Leah Tesfaye, Lauren Murphy"
output: pdf_document
date: "2025-04-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(reshape2)
library(DescTools)
library(gridExtra)
```


```{r}
df <- read.csv("C:\\Users\\ual-laptop\\Downloads\\Popular_Spotify_Songs.csv")

head(df)
```



```{r}
# Another dataset was created, grouping song counts by artists so as to
# consolidate our data to the necessary columns.
artist_counting <- df %>%
  group_by(artist.s._name) %>%
  summarise(song_count = n(), .groups = 'drop') %>%
  arrange(desc(song_count))

top_artists <- artist_counting %>%
  top_n(10, song_count)

head(top_artists)
```
# Question 1: Who are the top artists with the most songs?

**Why?**
- Considering the large scale of data(1930-2023), we wanted to consolidate the data to see if the top artists would be modern(ex. Taylor Swift, The Weeknd) or older in terms of popularity(ex. Madonna, Michael Jackson)

**Hypothesis**
-The chart will be comprised of modern artists, as older artists may have lower appearances in the dataset, despite their fame. This could be due to the fact that older artists/groups were unable to take advantage of streaming services like Spotify at the time of peak popularity.

```{r}
plot1 <- ggplot(top_artists, aes(x = reorder(artist.s._name, -song_count), 
                y = song_count, fill = reorder(artist.s._name, -song_count))) +
  geom_bar(stat = "identity") +
  labs(title = "Top 10 Artists with the Most Songs",
       x = "Artist",
       y = "Number of Songs",
       fill = "Artist name") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
  
print(plot1)
```
One of our key findings was that a few artists like Taylor Swift and The Weeknd were streamed the most compared to the other artists on the chart. 
Another observation was that the majority of top artists are English-speaking, but 3 groups are not(BTS, Bad Bunny, Feid). This represents the growing influence of non-English music & artists on the global music history.


# Question 2: How are song count and avg. valence related?

**Why?**
- Looking at song count in relation to another variable, such as valence(happiness) or liveness, can help us better understand potential relationships and the reasoning for different behaviors throughout the year.

**Hypothesis**
- May-July and December will have higher song counts, with drops occurring in August-October. In terms of valence, the trend will follow a similar path as song count.

```{r}
# Filters the data using 2022, grouping the data by month. The dataset also
# holds song count and average valence by month.
combined_data <- df %>%
  filter(released_year == 2022) %>%
  group_by(released_month) %>%
  summarise(
    song_count = n(),
    avg_energy = mean(valence_., na.rm = TRUE),
    .groups = "drop"
  )

# Song count plot. The song and energy plots were separated due to the different
# scaling along the Y-axis. We believed it would be easier to look at the 2 
# plots vertically, with the months lined up.

plot_song <- ggplot(combined_data, aes(x = released_month, y = song_count)) +
  geom_line(color = "lightblue", size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = "Song Count by Month (2022)",
    x = "Month",
    y = "Number of Songs"
  ) +
  theme_minimal()

# Energy plot
plot_energy <- ggplot(combined_data, aes(x = released_month, y = avg_energy)) +
  geom_line(color = "pink", size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    title = "Average Valence by Month (2022)",
    x = "Month",
    y = "Average Valence"
  ) +
  theme_minimal()


```

```{r}
# Stacking the two line graphs vertically for easier comparison
grid.arrange(plot_song, plot_energy, ncol = 1)
```

- One of our key findings was that months like May and December tend to have the most releases, in comparison to months with higher song valence being March, May-August and December. 

- One reason for the higher song count and average valence values in May could be due to seasonal changes, with artists and record labels releasing upbeat, happy music in anticipation of the summer season. 

- There are also points of conflict, with a peak in song count and drop in average valence in October. Moodier, more introspective music is often released during the fall months, which can explain the drop in valence. However, this does not necessarily correlate to lower song counts. Taylor Swift's Midnights album was released in October of 2022, with the music a mix of retro-pop and deeper, more introspecive tracks.

- Overall, we were able to better understand the potential relationship between valence and song releases throughout the year, as well as the potential reasons behind conflicting observations.
